The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-06-13 manuscript/5pillars.Rmd
#18 fixing some details

Computational reproducibility is a simple premise in theory, but in practice it is difficult to
achieve.
Building upon past efforts and proposals to maximize reproducibility and rigor in
bioinformatics, we present a framework called the five pillars of reproducible computational
research.
These include (1) code version control and sharing, (2) compute environment control,
(3) literate programming, (4) documentation and (5) FAIR data sharing.
These best practices will ensure that computational research work can be
reproduced quickly and easily, long into the future.
This guide is designed for bioinformatic data analysts, but should be relevant for other
domains.

@@

Around the same time, Project Jupyter was developed to provide a "notebook" type interface, incorporating
R, Python, Julia and other computer language chunks into documents [@kluyver2016; @granger2021; @Perkel2018-rz].
A noteworthy attempt to use literate programming in Jupyter notebooks to aim for reproducibility 
has been the analysis of RNA-seq samples from patients infected by the Zika virus [@wang2016].
In this work, authors provided along the traditionnal research paper a 
[Jupyter notebook](https://nbviewer.org/github/maayanlab/Zika-RNAseq-Pipeline/blob/master/Zika.ipynb) 
that performed and documented the entire analysis procedure.
The MyST-NB (short for Markedly Structured Text notebook) is built on top of Jupyter notebooks or Markdown documents
and has some notable advantages around flexibility of content that make it an attractive choice for executable
article authoring [@DuPre2022-bk].

@@

Although DVCSs assist with code sharing, they are not considered a long-term archiving solution,
as recalled by the complete shutdown of the Google Code platform in 2015
and the end of Mercurial support by the Bitbucket platform in 2020.
The code needs to be deposited to a long-term repository.
[Zenodo](https://zenodo.org/) and [Figshare](https://figshare.com/) are often used
to store the content of a repository but do not retain the history of the development of the source code.
[Software Heritage](https://www.softwareheritage.org/) [@dicosmo2017; abramatic2018] is the universal software archive
whose mission is to collect, preserve and share source code. Software Heritage provides
an intrinsic persistent identifier called SWHID that allows to cite the archived source code
in the respective journal article/preprint [@cosmo2020]. The archival of the source code in Software Heritage
can be performed [manually](https://archive.softwareheritage.org/save/)
or automated by a [webhook](https://docs.softwareheritage.org/user/save_code_now/webhooks/index.html)
within continuous integration.

@@

#border
polygon(c(0,12,12,0),c(0,0,8,8),border="gray")

@@

A lack of technical computer skills in many research groups is another culprit.
Specific workshop-based training in these areas could help, but the most scalable solution would be to
provide comprehensive beginner-friendly step-by-step guides for frequently used routines.
For example, our group has just released such a protocol that walks novice users through the process
of implementing functional enrichment analysis of gene expression data with the five pillars principles
@Ziemann2023-nm.
More such protocols are needed to make extreme reproducibility mainstream.
A compendium of publicly available learning materials around the five pillar concepts are
provided in the supplementary information.

>>>>

Computational reproducibility is a simple premise in theory, but is difficult to
achieve in practice.
Building upon past efforts and proposals to maximize reproducibility and rigor in
bioinformatics, we present a framework called the five pillars of reproducible computational
research.
These include (1) code version control and sharing, (2) compute environment control,
(3) literate programming, (4) documentation and (5) FAIR data sharing.
These best practices will ensure that computational research work can be
reproduced quickly and easily, long into the future.
This guide is designed for bioinformatic data analysts, but should be relevant for other
domains.

@@

Around the same time, Project Jupyter was developed to provide a "notebook" type interface, incorporating
R, Python, Julia and other computer language chunks into documents [@kluyver2016; @granger2021; @Perkel2018-rz].
A noteworthy attempt to use literate programming in Jupyter notebooks to aim for reproducibility
has been the analysis of RNA-seq samples from patients infected by the Zika virus [@wang2016].
In this work, authors provided along the traditional research paper a
[Jupyter notebook](https://nbviewer.org/github/maayanlab/Zika-RNAseq-Pipeline/blob/master/Zika.ipynb)
that performed and documented the entire analysis procedure.
The MyST-NB (short for Markedly Structured Text notebook) is built on top of Jupyter notebooks or Markdown documents
and has some notable advantages around flexibility of content that make it an attractive choice for executable
article authoring [@DuPre2022-bk].

@@

Although DVCSs assist with code sharing, they are not considered a long-term archiving solution,
as recalled by the complete shutdown of the Google Code platform in 2015
and the end of Mercurial support by the Bitbucket platform in 2020.
The code needs to be deposited to a long-term repository.
[Zenodo](https://zenodo.org/) and [Figshare](https://figshare.com/) are often used
to store the content of a repository but do not retain the history of the development of the source code.
[Software Heritage](https://www.softwareheritage.org/) [@dicosmo2017; @abramatic2018] is the universal
software archive whose mission is to collect, preserve and share source code.
Software Heritage provides an intrinsic persistent identifier called SWHID that allows to cite the
archived source code in the respective journal article/preprint [@cosmo2020].
Archiving source code in Software Heritage can be performed
[manually](https://archive.softwareheritage.org/save/) or automatically using a
[webhook](https://docs.softwareheritage.org/user/save_code_now/webhooks/index.html) within continuous
integration.

@@

@@

A lack of technical computer skills in many research groups is another culprit.
Specific workshop-based training in these areas could help [@Williams2023-ic].
The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
[@Wilson2014-mh].
However, the most scalable solution would be to provide comprehensive beginner-friendly step-by-step
guides for frequently used routines.
For example, our group has just released such a protocol that walks novice users through the process
of implementing functional enrichment analysis of gene expression data with the five pillars principles
@Ziemann2023-nm.
More such protocols are needed to make extreme reproducibility mainstream.
A compendium of publicly available learning materials around the five pillar concepts are
provided in the supplementary information.
